# 4. Modelling

First of all, we will build different model based on the population sample in every state of 2018, the election result of 2016, and the people shoot from 2013 to 2020. 

The election result of 2016 is right in the middle of the period took for people shoot and we assume the population do not vary a lot between 2016 and 2018.

**Variable**

Rate..All.People. : Rate of peaople killed by 125'000 inhabitants (2013-2020)

Black.White.Disparity : Disparity in rate between black and white (2013-2018)

percent_democrat_2016 : percent of democrat  for 2016

percent_black : percent of black in 2018 

percent_asian : percent of asian in 2018 

percent_white : percent of white in 2018

percent_hispanic : percent of hispanic in 2018

violent_crime_2016 : Violent crime rates per 100000 inhabitants in 2016

Total.Population : Total population in 2018

```{r, echo = FALSE, message = FALSE, warning=FALSE}
rp <- read.csv(file = here::here("Data/resident_population.csv"), sep = ";", fileEncoding = 'UTF-8-BOM')
Gender_Composition <- read.csv(file = here::here("Data/Gender_Composition.csv"))
pk <- read.csv(file = here::here("Data/2013_2021_police_killings.csv"), sep = ";")
Killingsbystate <- read.csv(file = here::here("Data/2013-2021_killings_by_state.csv"), sep = ";", fileEncoding = 'UTF-8-BOM')
Race_and_ethnicity <- read.csv(file = here::here("Data/Race_and_ethnicity.csv"))
democratic_vs_republican_votes_by_usa_state_2020 <- read.csv(file = here::here("Data/democratic_vs_republican_votes_by_usa_state_2020.csv"))
democratic_vs_republican_votes_usa_2016 <- read.csv(file = here::here("Data/democratic_vs_republican_votes_usa_2016.csv"))
rp <- read.csv(file = here::here("Data/resident_population.csv"), sep = ";", fileEncoding = 'UTF-8-BOM')
cr <- read.csv(file = here::here("Data/crime_rate.csv"), sep = ";")


election <- merge(democratic_vs_republican_votes_by_usa_state_2020, democratic_vs_republican_votes_usa_2016, by = "state")
election <- election %>% 
  rename(
    percent_democrat_2016 = percD,
    percent_democrat_2020 = percent_democrat
  )
election$percent_democrat_2016 <- election$percent_democrat_2016 * 100
election_percent <- election %>% select(percent_democrat_2020, percent_democrat_2016, state)

names(Killingsbystate)[1]<-paste("state")
state <- Killingsbystate[c("state", "State.Abbreviation")]

pk_count_armed <- count(pk, Victim.s.race, Armed.Unarmed.Status)

#subset(subset(pk_count_armed, Victim.s.race == "White"))[2:3][2:6,][-4,]
#subset(subset(pk_count_armed, Victim.s.race == "White"),Armed.Unarmed.Status== "Allegedly Armed") [3]
#subset(subset(pk_count_armed, Victim.s.race == "White"),Armed.Unarmed.Status== "Unarmed/Did Not Have Actual Weapon")[3]
#subset(subset(pk_count_armed, Victim.s.race == "Black"))[2:3][2:5,]
#subset(subset(pk_count_armed, Victim.s.race == "Black"),Armed.Unarmed.Status== "Allegedly Armed")[3]
#subset(subset(pk_count_armed, Victim.s.race == "Black"),Armed.Unarmed.Status== "Unarmed/Did Not Have Actual Weapon")[3]

#plt_white <- subset(subset(pk_count_armed, Victim.s.race == "White"))[2:3][2:6,][-4,]
#ggplotwhite <- ggplot(plt_white, aes(x=Armed.Unarmed.Status, y=n)) + 
 # geom_bar(stat="identity", fill="White") +
#  labs(title = "White Armed/unaremed Status",
  #     subtitle = "2013-2020 ",
  #     x = "Armed Status",
  #     y= "Number") +
 #   geom_text(aes(label=n)) 
#ggplotwhite

#plt_black <- subset(subset(pk_count_armed, Victim.s.race == "Black"))[2:3][2:5,]
#ggplotblack <- ggplot(plt_black, aes(x=Armed.Unarmed.Status, y=n)) + 
#  geom_bar(stat="identity", fill="Black") +
#  labs(title = "Black Armed/unaremed Status",
 #      subtitle = "2013-2020 ",
#       x = "Armed Status",
 #      y= "Number") +
#    geom_text(aes(label=n), color = "white") 
#ggplotblack

#grid.arrange(ggplotwhite, ggplotblack, ncol=2)

```


## Can we predict the number of people getting shot by the police with the democratic percentage?

## Correlation table
```{r, echo = FALSE, message = FALSE, warning=FALSE}
#setp data
election <- merge(democratic_vs_republican_votes_by_usa_state_2020, democratic_vs_republican_votes_usa_2016, by = "state")
Killingsbystate$X..White <- Killingsbystate$'White.Population' / Killingsbystate$'Total.Population'
Killingsbystate$X..Hispanic <- Killingsbystate$'Hispanic.Population' / Killingsbystate$'Total.Population'
Killingsbystate$X..Asian <- Killingsbystate$'Asian.Population' / Killingsbystate$'Total.Population'

b_pop <- data.frame(Killingsbystate[c("state","Rate..All.People.", "X..People.Killed", "Total.Population", "X..African.American","X..White", "X..Hispanic", "X..Asian", "X..Victims.Black","Rate..Black.People.", "Disparity.in.Rate", "X..White", "Disparity", "Disparity.in.Rate", "Black.White.Disparity")])
b_pop <- b_pop[1:51,]
vote_rel <- merge(b_pop, election_percent, by = "state")

vote_rel$X..African.American <- as.numeric(sub(",", ".", vote_rel$X..African.American, fixed = TRUE))
vote_rel$Disparity <- as.numeric(sub(",", ".", vote_rel$Disparity, fixed = TRUE))
vote_rel$X..Victims.Black <- as.numeric(sub(",", ".", vote_rel$X..Victims.Black, fixed = TRUE))
vote_rel$Rate..All.People. <- as.numeric(sub(",", ".", vote_rel$Rate..All.People., fixed = TRUE))
vote_rel$Rate..Black.People. <- as.numeric(sub(",", ".", vote_rel$Rate..Black.People., fixed = TRUE))
vote_rel$Black.White.Disparity <- as.numeric(sub(",", ".", vote_rel$Black.White.Disparity, fixed = TRUE))
vote_rel$Disparity.in.Rate <- as.numeric(sub(",", ".", vote_rel$Disparity.in.Rate, fixed = TRUE))

vote_rel$X..Victims.Black <- vote_rel$X..Victims.Black * 100
vote_rel$Disparity <- vote_rel$Disparity * 100
vote_rel$X..African.American <- vote_rel$X..African.American * 100
vote_rel$X..White <- vote_rel$X..White * 100
vote_rel$X..Asian <- vote_rel$X..Asian * 100
vote_rel$X..Hispanic <- vote_rel$X..Hispanic * 100
vote_rel_new <- merge(vote_rel, cr[c("state", "X2016")], by = "state")
vote_rel_new <- vote_rel_new[c("Rate..All.People.", "Black.White.Disparity", "percent_democrat_2016", "X2016", "X..African.American", "X..White", "X..Hispanic", "X..Asian", "Total.Population")]

vote_rel_new <- vote_rel_new %>% 
  rename(
    percent_black = X..African.American,
    percent_asian = X..Asian,
    percent_white = X..White,
    percent_hispanic = X..Hispanic,
    violent_crime_2016 = X2016
  )

#$disparity_diff <- vote_rel$X..Victims.Black / vote_rel$X..African.American

par(mfrow=c(1,1))
#vote_no_state <- vote_rel[,-(1)]
c <- cor(vote_rel_new)
#corrplot(c, method = "number")

palette = colorRampPalette(c("green", "white", "red")) (20)
ggcorr(c, nbreaks = 8, palette = "RdGy", label = TRUE, label_size = 3, label_color = "white", legend.size = 9, digits = 3)

#heatmap(x = c, col = palette, symm = TRUE)

vote_rel_new_no8 <- vote_rel_new[vote_rel_new$Disparity.in.Rate <8, ]
vote_rel_no45 <- vote_rel[vote_rel$X..People.Killed >45, ]
```


Rate all people is strongly correlated with violent crime 2016. This does make sense, more violent states lead to more police shooting. It is also negatively correlated with percent democrat 2016



Percent_democrat_2016 is slightly negativel correlated with violent_crime_2016. 

Percent of white is srtrongly negatively correlated to percent_Democrat while percent black is positevely correlated. Black population vote more for democrat president than thre white population who prefer republican president. 


## Can we predict the number of people getting shot by the police with the democratic percentage?

### Plot 1
```{r, echo = FALSE, message = FALSE, warning=FALSE, results='hide'}

#Rate..All.People
ggplot(vote_rel_new, aes(x=percent_democrat_2016, y=Rate..All.People.)) + 
  geom_point(shape=18, color="blue")+
  geom_smooth(method=lm,  linetype="dashed",
             color="darkred", fill="blue") +
  ggtitle("Rate..All.People - percent_democrat_2016") +
  xlab("Percent of democrat in 2016 (percent_democrat_2016)") + ylab("Number of people shoot by 125'000 habitant ( Rate..All.People.)")

```

First of all, there is a clear negative correlation between the percentage of democrat in a stat and the number of people killed by 125'000 people. Police in more democratic state seems to kill less people every year


### Plot 2
```{r, echo = FALSE, message = FALSE, warning=FALSE}
ggplot(vote_rel_new, aes(x=violent_crime_2016, y=Rate..All.People.)) + 
  geom_point(shape=18, color="blue")+
  geom_smooth(method=lm,  linetype="dashed",
             color="darkred", fill="blue") +
  ggtitle("Rate..All.People - Violent_crime 2016") +
  xlab("Violent Crime Rates per 100'000 habitants in 2016 (violent_crime_2016)") + ylab("Number of people shoot by 125'000 habitant")
```

Secondly, Violent Crime Ratest is strongly correlated to the number of people killed. Obviously, more violent crimes in a state lead to more shooting by the police.


### Multilinear regression
```{r, echo = FALSE, message = FALSE, warning=FALSE}
lm.fit <- lm(Rate..All.People. ~ percent_democrat_2016 + violent_crime_2016 + percent_black + percent_white + percent_hispanic + percent_asian + Total.Population, data=vote_rel_new)
tab_model(lm.fit)

#knitr::kable(bptest(lm.fit)[4])
```

An R-squared of 0.709 is very high as well as the adjusted R-squared. 

Percent white and percent black are extremely significative (p<0.001).Percent asian is also significative with 2 stars (p<0.01) and in the end percent hispanic but with only 1 star (p<0.5).

Weirdly, all demographic variable are negative.

The variable we were interested primarily, percent democrat 2016, is also significative but with p<0.05. In this regression, police in less democratic state definitely shoot more people during arrest than more democratic state. 

As expected, the Total Population variable has no influence on Rate All People. Maybe we can remove this variable to get a better regression.

### Regression Diagnosis
```{r, echo = FALSE, message = FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(lm.fit)
par(mfrow=c(1,1))

```
**Residual vs Fitted :** The curve is flat meaning there is no heteroscedasticity 

**Normal Q-Q :** The linearity regression fit perfectly all the variable. The curve stay flat on both extreme. 

**Scale-Location :** The red line is roughly horizontal across the plot; homoscedasticity is satisfied. 

**Residual vs Leverage :** Observation 5 is slightly out of the red line meaning this observation has a huge impact on our regression. Globally, there is no big anomaly.

The 4 plots indicate our multilinear regression is really good and there is no issue with it. 

### Exhaustive Search
```{r, echo = FALSE, message = FALSE, warning=FALSE}
regsubsets.out <-
regsubsets(Rate..All.People. ~ percent_democrat_2016 + violent_crime_2016 + percent_black + percent_white + percent_hispanic + percent_asian + Total.Population, data=vote_rel_new,
             nbest = 1,       # 1 best model for each number of predictors
             nvmax = NULL,    # NULL for no limit on number of variables
             force.in = NULL, force.out = NULL,
             method = "exhaustive")
plot(regsubsets.out, scale = "adj", main = "Adjusted-r2")


```

Three models have a close adj R2; 5, 6 or 7 variables. Let's take a deeper look and choose the best one according to other criteria.


### Forward selection
```{r, echo = FALSE, message = FALSE, warning=FALSE}
par(mfrow = c(1, 1))
model.f2 <- regsubsets(Rate..All.People. ~ percent_democrat_2016 + violent_crime_2016 + percent_black + percent_white + percent_hispanic + percent_asian + Total.Population, data=vote_rel_new, nvmax = 20, method = "forward")
plot(model.f2)
sum.model.f <- summary(model.f2)
#sum.model.f
par(mfrow = c(1, 2))
plot(sum.model.f$cp, xlab = "No. of variables", ylab = "RSS", type = "l")
points(which.min(sum.model.f$cp), sum.model.f$cp[which.min(sum.model.f$cp)], col = "blue", cex = 3, pch = 20)
plot(sum.model.f$cp, xlab = "No. of variables", ylab = "Cp", type = "l")
points(which.min(sum.model.f$cp), sum.model.f$cp[which.min(sum.model.f$cp)], col = "blue", cex = 3, pch = 20)
plot(sum.model.f$bic, xlab = "No. of variables", ylab = "BIC", type = "l") #6 variables is the best model
points(which.min(sum.model.f$bic), sum.model.f$bic[which.min(sum.model.f$bic)], col = "blue", cex = 3, pch = 20)
plot(sum.model.f$adjr2, xlab = "No. of variables", ylab = "Adj. R2", type = "l")
points(which.max(sum.model.f$adjr2), sum.model.f$adjr2[which.max(sum.model.f$adjr2)], col = "blue", cex = 3, pch = 20)
#coef(model.f2,6)
```

All indicator suggest for the forward selection the model with all variables included (7 variables)

In the forward selection, percent_asian is surprisingly the last variable to be added to the regression and violent_crime_2016 is the first one as the plot suggested it. 

### Backward selection
```{r, echo = FALSE, message = FALSE, warning=FALSE}

par(mfrow = c(1, 1))
model.f1 <- regsubsets(Rate..All.People. ~ percent_democrat_2016 + violent_crime_2016 + percent_black + percent_white + percent_hispanic + percent_asian + Total.Population, data=vote_rel_new, nvmax = 20, method = "backward")
plot(model.f1)
sum.model.f <- summary(model.f1)
#sum.model.f
par(mfrow = c(1, 2))
plot(sum.model.f$cp, xlab = "No. of variables", ylab = "RSS", type = "l")
points(which.min(sum.model.f$cp), sum.model.f$cp[which.min(sum.model.f$cp)], col = "blue", cex = 3, pch = 20)
plot(sum.model.f$cp, xlab = "No. of variables", ylab = "Cp", type = "l")
points(which.min(sum.model.f$cp), sum.model.f$cp[which.min(sum.model.f$cp)], col = "blue", cex = 3, pch = 20)
plot(sum.model.f$bic, xlab = "No. of variables", ylab = "BIC", type = "l") #6 variables is the best model
points(which.min(sum.model.f$bic), sum.model.f$bic[which.min(sum.model.f$bic)], col = "blue", cex = 3, pch = 20)
plot(sum.model.f$adjr2, xlab = "No. of variables", ylab = "Adj. R2", type = "l")
points(which.max(sum.model.f$adjr2), sum.model.f$adjr2[which.max(sum.model.f$adjr2)], col = "blue", cex = 3, pch = 20)
#coef(model.f1,6)

```
However, for the backward selection, it is less explicit. The best model seems to be the one with 6 variables.

Obviously, Total.Population is the first variable to me removed
from the backward selection process. This variable is the less significative in the base model with the 7 variables. 


### Conclusion
To conclude, the best model is the one with all variables except Total.Population. Despite the adjusted R-squared being slightly higher with this variable, the other test proved the model is better without it.


## Can we explain the disparity between black and white shots with the democratic percentage of the state?


### Plot 1
```{r, echo = FALSE, message = FALSE, warning=FALSE}
par(mfrow = c(1, 1))
ggplot(vote_rel_new, aes(x=percent_democrat_2016, y=Black.White.Disparity)) + 
  geom_point(shape=18, color="blue")+
  geom_smooth(method=lm,  linetype="dashed", color="darkred", fill="blue") +
  ggtitle("Black.White.Disparity - percent_democrat_2016") +
  xlab("Percent of democrat in 2016") + ylab("More time black killed compared to white")

```


### Plot 2
```{r, echo = FALSE, message = FALSE, warning=FALSE}
par(mfrow = c(1, 1))
ggplot(vote_rel_new, aes(x=violent_crime_2016, y=Black.White.Disparity)) + 
  geom_point(shape=18, color="blue")+
  geom_smooth(method=lm,  linetype="dashed", color="darkred", fill="blue") +
  ggtitle("Black.White.Disparity - violent_crime_2016") +
  xlab("Violent crime rate per 100'000 habitans in 2016)") + ylab("More time black killed compared to white")

```

### Plot 3
```{r, echo = FALSE, message = FALSE, warning=FALSE}
par(mfrow = c(1, 1))
ggplot(vote_rel_new, aes(x=percent_black, y=Black.White.Disparity)) + 
  geom_point(shape=18, color="blue")+
  geom_smooth(method=lm,  linetype="dashed", color="darkred", fill="blue") +
  ggtitle("Black.White.Disparity - percent_black") +
  xlab("Percent of black") + ylab("More time black killed compared to white")

```

### Multilinear regression
```{r, echo = FALSE, message = FALSE, warning=FALSE}
lm.fit <- lm(Black.White.Disparity ~ percent_democrat_2016 + violent_crime_2016 + percent_black + percent_white + percent_hispanic + percent_asian + Total.Population, data=vote_rel_new)
tab_model(lm.fit)
```

An R-squared of 0.124 is very low and a negative adjusted R-squared is really bad.

On top of that, none of the variable are significative. 

Basically, this multilinear regression do not predict or explain Rate all People and there is no need to pick a "bette" model among this one. 

### Regression Diagnosis
```{r, echo = FALSE, message = FALSE, warning=FALSE}
par(mfrow=c(2,2))
plot(lm.fit)
par(mfrow=c(1,1))

```
**Residual vs Fitted :** The curve is not flat. This show us there is no homoscedasticity.

**Normal Q-Q :** The linearity regression do not fit observation 44 and 39 that are clearly way above the line.

**Scale-Location :** The red line has a banana shape confirming what said above: there is heteroscedasticity. 

**Residual vs Leverage :** Observation 11 is out of the red ine and observation 39 and 2 are close to it. Thi regression suffer from this variable that have a huge influence on it. 

The 4 plots indicate our multilinear regression is not a good fit at all.

