---
title: "Does race and politics have an impact on police violence in the United States ?"
author: "Jérémy Olivier and Chloé Mousse"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r, echo = FALSE, message = FALSE, warning=FALSE}
#Dowload data
Gender_Composition <- read.csv(file = here::here("Data/Gender_Composition.csv"))
democratic_vs_republican_votes_by_usa_state_2020 <- read.csv(file = here::here("Data/democratic_vs_republican_votes_by_usa_state_2020.csv"))
democratic_vs_republican_votes_usa_2016 <- read.csv(file = here::here("Data/democratic_vs_republican_votes_usa_2016.csv"))
Race_and_ethnicity <- read.csv(file = here::here("Data/Race_and_ethnicity.csv"))
offender_sex_and_race <- read.csv(file = here::here("Data/offender_sex_and_race.csv"), fileEncoding = 'UTF-8-BOM')
rp <- read.csv(file = here::here("Data/resident_population.csv"), sep = ";", fileEncoding = 'UTF-8-BOM')
Killingsbystate <- read_delim(file = here::here("Data/2013-2021_killings_by_state.csv"), 
     delim = ";", escape_double = FALSE, trim_ws = TRUE)
pk<- read_delim(file = here::here("Data/2013_2021_police_killings.csv"), 
     delim = ";", escape_double = FALSE, trim_ws = TRUE)

```

## Police composition

### Table 1 : Gender composition

#### 1) Cleaning the data set 

```{r, echo = FALSE, message = FALSE, warning=FALSE}
Gender_Composition <- select(Gender_Composition,c(2,4:7,11))
Gender_Composition$share <- Gender_Composition$share * 100 

kable(Gender_Composition) %>% 
  kable_styling(fixed_thead = T, full_width = F, position = "left") %>% 
  row_spec(0, bold = T, italic = T, color = "white", background = "darkslategray") %>% 
  scroll_box(height = "400px", width = "75%")

```

#### 2) Data vizualisation

In the united state in 2020, there is roughly 51% of women and the black population represent approximately 13,5% of the total population. 

```{r, echo = FALSE, message = FALSE, warning=FALSE}

Gender_Composition[Gender_Composition$Gender == "Female", ]   %>%
  tail(10) %>%
  ggplot( aes(x=Year, y=share)) +
  geom_line() +
  geom_point() +
  expand_limits(x = c(2013, 2020)) +
  ggtitle("Women percentage in the police \n 2014-2019") +
  xlab("Year") + ylab("PErcentage")

#head(Race_and_ethnicity)
#summary(Race_and_ethnicity)
#str(Race_and_ethnicity)
#colSums(is.na(Race_and_ethnicity))
```

On the opposite, the number of women in the police is really low. INTRODUIRE DONNEE SUR LA POPULATION TOTALE HOMME FEMME POUR COMPARER ? 
There has always been more women than men in the United States. Women represent about 51% against 49% of men. On the graph above we clearly see that women are only 15% of the total police workforce, still in 2020. 

### Table 2 : Race and ethnicity composition

#### 1) Cleaning the data set

```{r, echo = FALSE, message = FALSE, warning=FALSE}
Race_and_ethnicity$share <- round(Race_and_ethnicity$share * 100,2)

RAE <- Race_and_ethnicity %>%
  subset(.,Race=="Black" |  Race=="White") %>%
  subset(.,Slug.Geography=="united-states") %>%
  select(.,c(2,4,6,9:11,16))

kable(RAE) %>% 
  kable_styling(fixed_thead = T, full_width = F, position = "left") %>% 
  row_spec(0, bold = T, italic = T, color = "white", background = "darkslategray") %>% 
  scroll_box(height = "400px", width = "95%")
```

#### 2) Data vizualisation

```{r, echo = FALSE, message = FALSE, warning=FALSE}

Race_and_ethnicity_b <- Race_and_ethnicity[Race_and_ethnicity$Race == "Black",]
Race_and_ethnicity_c <- Race_and_ethnicity_b[Race_and_ethnicity_b$Slug.Geography == "united-states",]

bp2 <- Race_and_ethnicity_c[Race_and_ethnicity_c$ID.PUMS.Ethnicity.Parent == "Not Hispanic" ,]   %>%
  tail(10) %>%
  ggplot( aes(x=Year, y=share)) +
  geom_line() +
  geom_point()+
  expand_limits(x = c(2013, 2020)) +
  ggtitle("Black percentage in the police \n 2014-2019") +
  xlab("Year") + ylab("Percentage")


#head(Gender_Composition)
#summary(Gender_Composition)
#str(Gender_Composition)
#colSums(is.na(Gender_Composition))


rp$total_population <- rp$White + rp$Black.or.African.American + rp$American.Indian.and.Alaska.Native+ rp$Asian + rp$Native.Hawaiian.and.Other.Pacific.Islander + rp$Two.or.more.races
#str(rp)
rp$black_share <- (rp$Black.or.African.American / rp$total_population) * 100


bp1 <- rp %>%
  #tail(25) %>%
  ggplot( aes(x=Characteristic, y=black_share)) +
  geom_line() +
  geom_point()+
  #expand_limits(x = c(2016, 2020)) +
  xlim(2013, 2020) +
  ggtitle("Black percentage in the USA \n 2014-2019") +
  xlab("Year") + ylab("Percentage")

grid.arrange(bp1, bp2, ncol=2)

```

For the past 20 years, the percentage of black population is rising in the united state and this trend is even stronger in the force police. However, the share in the police is approximately 1.5 point down compared to the share in the united state. Though, We can still agree that the black population is correctly represent in the police.


## All violent crime offenders 

#### 1) Cleaning and creating a new table

```{r, echo = FALSE, message = FALSE, warning=FALSE}
osr <- offender_sex_and_race
osr$offender_white <- round(osr$offender_white/osr$offender_total*100,2)
osr$offender_black <- round(osr$offender_black/osr$offender_total*100,2)
osr$offender_male <- round(osr$offender_male/osr$offender_total*100,2)
osr$offender_female <- round(osr$offender_female/osr$offender_total*100,2)

kable(osr) %>% 
  kable_styling(fixed_thead = T, full_width = F, position = "left") %>% 
  row_spec(0, bold = T, italic = T, color = "white", background = "darkslategray") 

```

In 2020, there were 538,203 violent-crime incident compared to only  276,363 in 2013. Violence is rising in the United Sates, but what population is more responsible of it ? 

```{r, echo = FALSE, message = FALSE, warning=FALSE}
#head(offender_sex_and_race)
#summary(offender_sex_and_race)
#str(offender_sex_and_race)
#colSums(is.na(offender_sex_and_race))

offender_sex_and_race$other_race <- offender_sex_and_race$offender_total - offender_sex_and_race$offender_black - offender_sex_and_race$offender_white
offender_sex_and_race$other_sex <- offender_sex_and_race - offender_sex_and_race$offender_female - offender_sex_and_race$offender_male
offender_sex_and_race$share_race <- offender_sex_and_race$offender_black / offender_sex_and_race$offender_total
offender_sex_and_race$share_sex <- offender_sex_and_race$offender_female / offender_sex_and_race$offender_total
offender_sex_and_race$share_race_white <- offender_sex_and_race$offender_white / offender_sex_and_race$offender_total

offender_sex_and_race$share_race_white <- offender_sex_and_race$share_race_white * 100
offender_sex_and_race$share_race <-offender_sex_and_race$share_race*100
offender_sex_and_race$share_sex <- offender_sex_and_race$share_sex*100


#offender_sex_and_race   %>%
  #tail(10) %>%
  #ggplot( aes(x=year, y=share_race)) +
  #geom_line(data=data.frame(offender_sex_and_race$share_race_white),aes(color="Second line")) +
  #geom_line() +
  #geom_point() +
  #ggtitle("Black percentage in the offender \n 2013-2020") +
  #xlab("Year") + ylab("Percentage")

offender_sex_and_race_a <- data.frame(offender_sex_and_race[, c("year", "share_race")])
offender_sex_and_race_b <- data.frame(offender_sex_and_race[, c("year", "share_race_white")])
offender_sex_and_race_b <- offender_sex_and_race_b %>% 
  rename(
    share_race = share_race_white,
  )

ggplot(offender_sex_and_race_a, aes(year, share_race)) +
geom_line(aes(color="Black")) +
geom_line(data=offender_sex_and_race_b, aes(color="White")) +
geom_point(data=offender_sex_and_race_b) +
geom_point() +
scale_color_manual(values=c('#999999','#FFFFFF')) +
ggtitle("Black and white percentage in the offender \n 2013-2020") +
xlab("Year") + ylab("Percentage") 
``` 

First of all, despite represent only 13.5 percent of the population, black people represent more than 40 percent of all violent crime in the United States. However, for the past years, the share of black people involved in violent crime is going down except in 2019. 

```{r, echo = FALSE, message = FALSE, warning=FALSE}

offender_sex_and_race   %>%
  tail(10) %>%
  ggplot( aes(x=year, y=share_sex)) +
  geom_line() +
  geom_point() +
  ggtitle("Women percentage in the offender \n 2013-2020") +
  xlab("Year") + ylab("Percentage")

```

On the opposite, the percentage  violent crime done by women is very low.


## USA Elections

#### 1) Cleaning and creating a new table
  
We create a new table that regroup all the result of the election of 2016 and 2020 by merging the previous one on the state code. This step done, we rename the columns to focus on the percentage of democrats in each state in the USA. 

```{r, echo = FALSE, message = FALSE, warning=FALSE}

#CREATE NEW DF 
election <- merge(democratic_vs_republican_votes_by_usa_state_2020, democratic_vs_republican_votes_usa_2016, by = "state")
election <- election %>% 
  rename(
    percent_democrat_2016 = percD,
    percent_democrat_2020 = percent_democrat
  )
election$percent_democrat_2016 <- election$percent_democrat_2016 * 100
election <- election %>% mutate(across(starts_with("percent_democrat"), round, 1))
```

```{r, echo = FALSE, message = FALSE, warning=FALSE}
election_percent <- election %>% 
  select(state, percent_democrat_2016, percent_democrat_2020)

kable(election_percent) %>% kable_styling(fixed_thead = T, full_width = F, position = "left") %>% 
  row_spec(0, bold = T, italic = T, color = "white", background = "darkslategray") %>% 
  column_spec(1, color = "white", background = "cadetblue") %>% 
  scroll_box(height = "350px", width = "57%")

```

Almost every state got more democrat vote in 2020 than in 2016. 

#### 2) Election map 

```{r, echo = FALSE, message = FALSE, warning=FALSE}
#USA MAP ELECTION
all_states <- map_data("state")
election$region <- tolower(election$usa_state)
totaldf <- merge(all_states, election, by = "region")
# switched to data.table to fix the cut up map issue
# getting sort by region then order 
totaldt <- as.data.table(totaldf)
setkey(totaldt, region, order)

options(scipen=999) #turn off scientific notation

e1 <- ggplot(data = totaldt, 
       aes(x = long, y = lat, group = group, fill = percent_democrat_2016)) +
  geom_polygon() + coord_map() +
  scale_fill_gradientn("", colours=brewer.pal(9, "RdYlBu")) + ggtitle("USA election  2016 \n (percent of democrat vote)") 

e2 <- ggplot(data = totaldt, 
       aes(x = long, y = lat, group = group, fill = percent_democrat_2020)) +
  geom_polygon() + coord_map() +
  scale_fill_gradientn("", colours=brewer.pal(9, "RdYlBu")) + ggtitle("USA election  2020 \n (percent of democrat vote)") 
grid.arrange(e1, e2, ncol=2)

```

As previously stated, more states turned into being democrats in 2020. We can observe this small difference on the 2 maps. 

#### 3) Graphic visualization

```{r, echo = FALSE, message = FALSE, warning=FALSE}
#USA BARPLOT ELECTION
plot1 <- ggplot(election, aes(y= reorder(usa_state_code, percent_democrat_2016), x=percent_democrat_2016)) + 
  geom_bar(stat="identity", width=.5, fill="darkslategray") + 
  geom_text(aes(label = percent_democrat_2016), hjust = -0.5, size = 3) + 
  labs(title = "Democrat percentage by state",
       subtitle = "2016 ",
       x = "Percent",
       y= "State code") +
  geom_vline(xintercept = 50, 
             color = "red", size=0.7) +
  expand_limits(x = c(0, 75))

plot2 <- ggplot(election, aes(y= reorder(usa_state_code, percent_democrat_2020), x=percent_democrat_2020)) + 
  geom_bar(stat="identity", width=.5, fill="darkslategray") + 
  geom_text(aes(label = percent_democrat_2020), hjust = -0.5, size = 3) + 
  labs(title = "Democrat percentage by state",
       subtitle = "2020 ",
       x = "Percent",
       y= "State code") +
  geom_vline(xintercept = 50, 
             color = "red", size=0.7) +
  expand_limits(x = c(0, 75))

grid.arrange(plot1, plot2, ncol=2)
```


In the election of 2020, there is more state above the 50 percent line than 4 years ago. This situation led to the election of a democrat president.
